Supporting Information for "Geometric models overestimate lake depth due to imperfect slope measurement"

### 2020-07-29 manuscript/agujournaltemplate.tex
rm in prep refs

We categorized lakes based on their cross-section shape and reservoir class. For cross-section shape, we categorized lakes as either convex or concave following the method of \citeA{hakansonLakeFormLake1977} by computing normalized lake depth-area relationships (i.e. hypsographic curves) and assigning class membership based on whether a lake’s curve falls above or below that of a simple straight-sided cone (Figure S3). We further classified lakes using data from \citeA{polus2020}, which uses the output of a machine learning algorithm to assign a probability to each lake as to whether it is a reservoir or a natural lake. For our purposes, we determined a lake to be a reservoir if the classification probability was 0.75 or greater. The \citeA{polus2020} data product defines reservoirs as any permanent waterbody that has a water control structure likely to significantly impact flow or pool water, beyond simply controlling water level. It makes no distinction between different dam types or dam heights.

Covariates used in random forest modeling (Table S1, Equation \ref{eq3}) for lake elevation, area, island area, perimeter, shoreline development, watershed to lake area ratio, and hydrologic subbasin (i.e. HUC4s), were obtained from the LAGOS-US LOCUS database \cite{smith2020}. One such measure, that of shoreline development, is a measure of lake perimeter shape defined as:

@@

One surprising finding with respect to the relationship between true and proxy geometry measures when examined by lake class was the fact that there was no greater difference between proxy and true distances in reservoirs compared to natural lakes. This is contrary to the idea that most reservoirs are drowned river valleys where the deepest point is close to the edge at the end of the reservoir (i.e. next to the dam) rather than in the center of the reservoir \cite{lanza1985interactions}. One possible explanation is that the reservoir classification data from \citeA{polus2020} uses a more general definition of a reservoir (i.e. any permanent waterbody that has a water control structure likely to significantly impact flow or pool water) compared to that of conventional classifications that are tied to specific dam types or dam heights. Another possible explanation is that conventional reservoir classifications are conceptually biased towards more southern areas with few natural lakes (Figure S2).

>>>>

We categorized lakes based on their cross-section shape and reservoir class. For cross-section shape, we categorized lakes as either convex or concave following the method of \citeA{hakansonLakeFormLake1977} by computing normalized lake depth-area relationships (i.e. hypsographic curves) and assigning class membership based on whether a lake’s curve falls above or below that of a simple straight-sided cone (Figure S3). We further classified lakes using the output of a machine learning algorithm to assign a probability to each lake as to whether it is a reservoir or a natural lake. For our purposes, we determined a lake to be a reservoir if the classification probability was 0.75 or greater. Our reservoir classification data defines reservoirs as any permanent waterbody that has a water control structure likely to significantly impact flow or pool water, beyond simply controlling water level. It makes no distinction between different dam types or dam heights.

Covariates used in random forest modeling (Table S1, Equation \ref{eq3}) for lake elevation, area, island area, perimeter, shoreline development, watershed to lake area ratio, and hydrologic subbasin (i.e. HUC4s), were obtained from the LAGOS-US LOCUS database. One such measure, that of shoreline development, is a measure of lake perimeter shape defined as:

@@

One surprising finding with respect to the relationship between true and proxy geometry measures when examined by lake class was the fact that there was no greater difference between proxy and true distances in reservoirs compared to natural lakes. This is contrary to the idea that most reservoirs are drowned river valleys where the deepest point is close to the edge at the end of the reservoir (i.e. next to the dam) rather than in the center of the reservoir \cite{lanza1985interactions}. One possible explanation is that our reservoir classification data uses a more general definition of a reservoir (i.e. any permanent waterbody that has a water control structure likely to significantly impact flow or pool water) compared to that of conventional classifications that are tied to specific dam types or dam heights. Another possible explanation is that conventional reservoir classifications are conceptually biased towards more southern areas with few natural lakes (Figure S2).

### 2020-07-29 manuscript/si_template_2019.tex
rm in prep refs

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
\bibliography{lagosdepth}
%
% no need to specify bibliographystyle
%
% Note that ALL references in this supporting information file must also be referenced in the primary manuscript
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% if you get an error about newblock being undefined, uncomment this line:
%\newcommand{\newblock}{}

@@

\begin{figure}
      \begin{center}\includegraphics[width=0.45\textwidth]{../figures/gg_effort-1}\end{center}
            \caption{Comparison between characteristics of lakes with bathymetry data against lakes with depth from other sources in the LAGOSUS-Depth product \cite{stachelek2020}. The distance to urban area metric is calculated using data from the 2018 US Census Urban and Rural Classification.}\label{figs7}
      \end{figure}

@@

\begin{figure}
      \begin{center}\includegraphics[width=0.73\textwidth]{../figures/lgnemanual-vs-bathy-depth-1}\end{center}
      \caption{Comparison between reported depth and depth extracted from bathymetry surfaces by US State where reported depths come from the LAGOSUS-Depth product \cite{stachelek2020}. For this figure, no reported depth values originated from the same source as its corresponding bathymetry-derived value.}\label{figs2}
      \end{figure}

@@

\begin{table}
\caption{Summary of lake characteristics for the present study (and for lakes in the contiguous United States from \citeA{stachelek2020}). Predictor variables for computing random forest offsets (Equation 2) are printed in bold face. Dashes (-) indicate an identical sample size among this study and that of the contiguous United States.} \label{table1}
\centering
\setlength\tabcolsep{1.5pt} % default value: 6pt
\begin{tabular}{lllll}
  \hline
  Variable & Median & Q25 & Q75 & n\\
  \hline
  Max depth (m) & 8.2 (7) & 4.6 (3.7) & 14 (12) & 4850 (17700)\\
  \textbf{Elevation (m)} & 300 (340) & 180 (210) & 400 (460) & 4850 (17700)\\
  \textbf{Area (ha)} & 55 (33) & 21 (11) & 140 (100) & 4850 (17700)\\
  \textbf{Island area (ha)} & 0 (0) & 0 (0) & 0.18 (0.076) & 4850 (17700)\\
  \textbf{Perimeter (m)} & 4400 (3500) & 2500 (1800) & 8100 (7300) & 4850 (17700)\\
  \textbf{Shoreline development} & 1.7 (1.7) & 1.4 (1.4) & 2.1 (2.2) & 4850 (17700)\\
  \textbf{Watershed-lake ratio} & 7.8 (10) & 3.9 (4.4) & 17 (28) & 4850 (17700)\\
  Distance to deepest point (m) & 180 (-) & 110 (-) & 290 (-) & 4850 (-)\\
  Distance to lake center (m) & 240 (-) & 160 (-) & 380 (-) & 4850 (-)\\
  In-lake slope (m/m) & 0.046 (-) & 0.024 (-) & 0.079 (-) & 4850 (-)\\
  Nearshore land slope (m/m) & 0.077 (-) & 0.051 (-) & 0.11 (-) & 4850 (-)\\
  \hline
  \end{tabular}
\end{table}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
% \bibliography{lagosdepth}
%
% no need to specify bibliographystyle
%
% Note that ALL references in this supporting information file must also be referenced in the primary manuscript
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% if you get an error about newblock being undefined, uncomment this line:
%\newcommand{\newblock}{}

@@

\begin{figure}
      \begin{center}\includegraphics[width=0.45\textwidth]{../figures/gg_effort-1}\end{center}
            \caption{Comparison between characteristics of lakes with bathymetry data against lakes with depth from other sources in the LAGOSUS-Depth product. The distance to urban area metric is calculated using data from the 2018 US Census Urban and Rural Classification.}\label{figs7}
      \end{figure}

@@

\begin{figure}
      \begin{center}\includegraphics[width=0.73\textwidth]{../figures/lgnemanual-vs-bathy-depth-1}\end{center}
      \caption{Comparison between reported depth and depth extracted from bathymetry surfaces by US State where reported depths come from the LAGOSUS-Depth product. For this figure, no reported depth values originated from the same source as its corresponding bathymetry-derived value.}\label{figs2}
      \end{figure}

@@

\begin{table}
\caption{Summary of lake characteristics for the present study (and for lakes in the contiguous United States). Predictor variables for computing random forest offsets (Equation 2) are printed in bold face. Dashes (-) indicate an identical sample size among this study and that of the contiguous United States.} \label{table1}
\centering
\setlength\tabcolsep{1.5pt} % default value: 6pt
\begin{tabular}{lllll}
  \hline
  Variable & Median & Q25 & Q75 & n\\
  \hline
  Max depth (m) & 8.2 (7) & 4.6 (3.7) & 14 (12) & 4850 (17700)\\
  \textbf{Elevation (m)} & 300 (340) & 180 (210) & 400 (460) & 4850 (17700)\\
  \textbf{Area (ha)} & 55 (33) & 21 (11) & 140 (100) & 4850 (17700)\\
  \textbf{Island area (ha)} & 0 (0) & 0 (0) & 0.18 (0.076) & 4850 (17700)\\
  \textbf{Perimeter (m)} & 4400 (3500) & 2500 (1800) & 8100 (7300) & 4850 (17700)\\
  \textbf{Shoreline development} & 1.7 (1.7) & 1.4 (1.4) & 2.1 (2.2) & 4850 (17700)\\
  \textbf{Watershed-lake ratio} & 7.8 (10) & 3.9 (4.4) & 17 (28) & 4850 (17700)\\
  Distance to deepest point (m) & 180 (-) & 110 (-) & 290 (-) & 4850 (-)\\
  Distance to lake center (m) & 240 (-) & 160 (-) & 380 (-) & 4850 (-)\\
  In-lake slope (m/m) & 0.046 (-) & 0.024 (-) & 0.079 (-) & 4850 (-)\\
  Nearshore land slope (m/m) & 0.077 (-) & 0.051 (-) & 0.11 (-) & 4850 (-)\\
  \hline
  \end{tabular}
\end{table}
